---
title: "How R Can Work with Databases!"
output:
  html_document:
    df_print: paged
---


## SQL to R Basics  
* There are multiple packages supporting connectivity between R and database management systems (**DBMS**)  
* Popular DBMS's include **SQL Server, SQLite, MySQL and Oracle**  
* R package **DBI** manages separate front-end and back-end interfaces between R and the DBMS. Applications use the front-end interface commands, while DBMS-specific tasks are handled by back-end drivers like RSQLite and RMySQL.  
* **dbplyr** is a package that helps dplyr work with SQL tables through a DBI connection  
```{r, warning=FALSE, message=FALSE}
library(DBI)
library(dplyr)
library(dbplyr)
library(wordcloud2)
```

## Let's explore some data!!

* Kaggle data set drawn from the Meetups.com API   
* https://www.kaggle.com/sirpunch/meetups-data-from-meetupcom/version/1  
* Contains data on Meetup members, groups, venues, topics and events from three cities: New York, San Francisco and Chicago  
* Data uses almost 1.5 GB on disk - not convenient to load into R memory  
* Stored locally in a MySQL database called mydb

## Create the database connection  
Check results in Connections pane
```{r, warning=FALSE, message=FALSE}
# "dbcon" connection configured in /etc/odbc.ini
con <- dbConnect(odbc::odbc(), "dbcon")

```

```{r}
# alternative to using ODBC configuration file: specify connection string
con2 <- dbConnect(odbc::odbc(), .connection_string="Driver={MySQL}; Uid=rebecca; Pwd=dbpass; Host=localhost; Port=3306; Database=mydb; Socket=/var/run/mysqld/mysqld.sock;")

```

### List the tables in the SQL database
```{r}
mydb.tables <- dbListTables(con)

mydb.tables
```

### Get the number of rows in each table
```{r}

for(table in mydb.tables){
  query <- paste0("Select count(*) from ", table)
  rows <- tbl(con, sql(query)) %>% collect()
  print(paste0(table, ": ", rows[[1]], " rows"))
}

```


### Get column names of the 'members' table
```{r}

dbListFields(con, "members")
dbListFields(con, "groups")

```

### Run 'SELECT' query and save the results as a data frame  
```{r}
venue.df <- dbGetQuery(con, "SELECT DISTINCT venue_name, city from venues;")
topics.df <- dbGetQuery(con, "SELECT DISTINCT topic_name from members_topics;")
```


## Use dplyr to create queries
* tbl() lets you connect to a table  

```{r}

members <- tbl(con, "members")
groups <- tbl(con, "groups")
show_query(members)

groups %>%
  filter(group_name %like% '%basenji%') %>%
  head(100)

```

### Run simple piped commands on your tbl() as you would a dataframe 
* dplyr functions that work on database backends include select, filter, group_by, join and summarise  
* To store results use collect()  
```{r}
members.by.state <- members %>%
  select(member_id, state) %>%
  group_by(state=='CA') %>%
  summarise(members=n()) %>%
  collect()

members.by.state
```

### Let's see that SQL query
```{r}
show_query(members %>%
             select(member_id, state) %>%
             group_by(state=='CA') %>%
             summarise(members=n()))
```

## Create a visualization
#### What do other data meetups call themselves?

```{r}
options(digits=2)
who <- groups %>%
  filter(group_name %like% '%data%') %>%
  select(word=who) %>%
  collect() %>%
  group_by(word) %>%
  summarise(n=n()) %>%
  mutate(freq=n/sum(n)) %>%
  select(-n)
  
wordcloud2(who, size=6)

```




```{r}

```



### Add R data frame to SQL database as new table
```{r}
topics <- read.csv('topics.csv', 
                   stringsAsFactors = FALSE)

dbWriteTable(con, "topics", topics)

# check for success
dbListTables(con)
```

### Now drop the table
```{r}
# delete the table
dbExecute(con, "DROP TABLE topics;")

# check for success
dbListTables(con)
```

